\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[T1]{fontenc}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{geometry}
\usepackage{graphicx}
\geometry{margin=1in}

\title{Larkos Training Loop and Inference Runner}
\author{Witold Warcho\l{}}
\date{June 2, 2026}

\lstset{
  basicstyle=\ttfamily\small,
  keywordstyle=\color{blue},
  commentstyle=\color{gray},
  stringstyle=\color{red},
  breaklines=true,
  captionpos=b
}

\begin{document}

\maketitle

\textbf{Last updated: \today}

\section{Overview}

The Larkos system consists of two major components:

\begin{enumerate}
  \item \textbf{TrainingLoop}: Full differentiable training with gradient
    computation, emotional state updates, and cognitive fusion
  \item \textbf{LarkosRunner}: Inference-only forward pass that loads
    trained weights and runs without gradient accumulation
\end{enumerate}

Both components share the same core modules: the LarkosModel,
\texttt{\_TemporalAttentionEncoder}, \texttt{\_NeuronGraphReasoner},
\texttt{\_FusionTransformerHead}, embedding networks, and text codec. The
key difference is that TrainingLoop computes gradients and updates weights,
while LarkosRunner only performs forward inference using frozen weights.

\section{Core Mathematical Components}

\subsection{Fourier Encoding}

The input $x \in \mathbb{R}^D$ is encoded using positional-style Fourier
features:

\[
  \phi(x) = \left[ \sin(2^0 \pi x), \sin(2^1 \pi x), \ldots,
    \sin(2^{E-1} \pi x), \cos(2^0 \pi x), \cos(2^1 \pi x), \ldots,
    \cos(2^{E-1} \pi x) \right]
\]

where $E$ is the number of encodings ($\texttt{FOURIER\_ENCODINGS}$). The
output has dimension $2 \cdot D \cdot E$.

\textbf{Rationale}: Fourier encodings help the model learn periodic and
frequency-dependent patterns. They provide stable gradients compared to
raw inputs and help with generalization.

\subsection{Online Normalization}

Two normalization strategies track running statistics:

\subsubsection{OnlineMinMax}

Per-dimension running min/max via exponential moving average (EMA):

\[
  m_t^{(i)} = m_{t-1}^{(i)} + \alpha (x_t^{(i)} - m_{t-1}^{(i)})
\]

where $\alpha = 0.02$ (momentum). The normalized output is:

\[
  \tilde{x}^{(i)} = \frac{x^{(i)} - m}{M - m + \epsilon} \cdot 2 - 1
\]

clamped to $[-3, 3]$.

\textbf{Rationale}: EMA-based min/max prevents early outliers from
permanently warping the normalization range. The system gradually forgets
extreme values and re-centers around the actual distribution.

\subsubsection{OnlineMeanStd}

Per-dimension running mean and variance via EMA:

\[
  \mu_t^{(i)} = \mu_{t-1}^{(i)} + \alpha (x_t^{(i)} - \mu_{t-1}^{(i)})
\]

\[
  \sigma_t^{(i)} = (1 - \alpha) \left( \sigma_{t-1}^{(i)} +
    \alpha (x_t^{(i)} - \mu_{t-1}^{(i)})^2 \right)
\]

Output normalization:

\[
  \hat{x}^{(i)} = \frac{x^{(i)} - \mu}{\sqrt{\sigma} + \epsilon_f}
\]

where $\epsilon_f = 0.1$ (std floor).

\textbf{Rationale}: The std floor prevents near-constant dimensions from
dividing by zero and blowing up. It keeps low-variance dimensions genuinely
quiet rather than amplified by numerical instability.

\subsection{Temporal Context Window and Encoder}

The Fourier-encoded input is padded into a temporal window of size
$w = \texttt{TEMPORAL\_WINDOW} = 4$, forming a sequence:

\[
  X_{\text{seq}} = [x_{t-w+1}, x_{t-w+2}, \ldots, x_t]
  \in \mathbb{R}^{w \times \texttt{FOURIER\_OUT\_DIM}}
\]

Earlier positions are zero-padded. The sequence is passed through
\texttt{\_TemporalAttentionEncoder} before flattening:

\[
  X_{\text{enc}} = \text{TransformerEncoder}
    \!\left(\text{LayerNorm}(X_{\text{seq}} + \mathbf{P}_{\text{pos}})\right)
  \in \mathbb{R}^{w \times \texttt{FOURIER\_OUT\_DIM}}
\]

\[
  x_{\text{temporal}} = \text{reshape}(X_{\text{enc}})
  \in \mathbb{R}^{w \cdot \texttt{FOURIER\_OUT\_DIM}}
\]

where $\mathbf{P}_{\text{pos}} \in \mathbb{R}^{w \times d}$ is a learned
positional embedding. The encoder uses $\texttt{TEMPORAL\_NHEAD}=4$ heads,
$\texttt{TEMPORAL\_LAYERS}=2$ layers, and $\texttt{TEMPORAL\_DIM\_FF}=64$.
The output shape is identical to a flat concatenation of the window frames,
so the downstream \texttt{LarkosModel} contract is unchanged.

\textbf{Rationale}: The encoder gives the model an explicit temporal axis
to attend over. The learned positional vector lets the model distinguish
the current frame from older frames. Without the encoder, temporal position
is only implicit in the slot offset of the flat vector.

\textbf{Gradient note}: \texttt{temporal\_encoder} is listed in
\texttt{DETACHED\_LABELS} because chain-rule attenuation through many
layers puts its gradient norm structurally orders of magnitude below the
fusion head. Dead-module and oscillation checks still apply; only the
head-vs-module balance ratio is skipped.

\subsection{LarkosModel Internals}

The LarkosModel maps the $\text{MODEL\_INPUT\_DIM}$-dim temporal
observation to a $\text{MAX\_NEURONS}$-dim neuron-output vector. It is
a transformer encoder over a token sequence built by quantising each
input dimension into a soft distribution over a learnable vocabulary,
optionally prepended with a context CLS token.

\subsubsection{NumericTokenizer}

Each input dimension $x^{(i)}$ is encoded as a soft assignment over $V
= \text{VOCAB\_SIZE}$ learnable prototypes $\{p_j\}_{j=1}^{V}$:

\[
  w_{ij} = \frac{\exp\big(-(x^{(i)} - p_j)^2\big)}
                 {\sum_{j'} \exp\big(-(x^{(i)} - p_{j'})^2\big)}
\]

The per-feature token embedding is then a weighted sum over the
embedding table $E \in \mathbb{R}^{V \times d_{\text{model}}}$:

\[
  t_i = \sum_{j=1}^{V} w_{ij}\, E_j \quad \in \mathbb{R}^{d_{\text{model}}}
\]

producing a sequence $(t_1, \ldots, t_D)$ that the encoder attends
over.

\subsubsection{EmbeddingProjector (Context Stream)}

When a context string $s$ is supplied, it is encoded with a frozen
pretrained sentence-transformer (MiniLM) and projected down to
$\text{PROJ\_DIM}$:

\[
  c = \text{LayerNorm}\big(W_{\text{proj}}\cdot \text{ST}(s)\big),
  \quad \text{ST}(\cdot)\ \text{frozen}.
\]

It is then projected up to $d_{\text{model}}$ and prepended as a
CLS-style token, so the encoder output at position $0$ is the summary
driving the linear head.

\textbf{Per-epoch cache}: The same $s$ is consumed roughly $15\times$
per epoch (MC samples, main forward, MAML inner steps, verifier).
EmbeddingProjector memoises the most-recent raw ST output keyed on
$s$, so $\text{ST}(\cdot)$ fires once per epoch rather than once per
forward pass.

\textbf{Shared ST under deepcopy}: A custom
\texttt{\_\_deepcopy\_\_} returns a clone that re-uses the frozen ST
module by reference. Without this, MAML's per-epoch \texttt{copy.\\
deepcopy(model)} would clone the $\sim 22$M frozen MiniLM parameters
on every inner update.

\subsection{Text Bridge (Debug Readout)}

The \texttt{\_TextCodec} renders the C-side cognitive vector as English
text. It is a \emph{one-way readout}, never trained: previous attempts
established that an 8-dim cognitive prefix cannot steer a frozen GPT-2,
and a learned text objective fought $L_{\text{base}}$.

\subsubsection{Source: Cognitive Vector, Not Head Output}

The bridge consumes the C-side $\text{fused\_cog\_raw} \in
\mathbb{R}^{\text{FUSION\_DIM}}$ vector (pre-transformer) rather than
the final $\mathbb{R}^{\text{MAX\_NEURONS}}$ head output. The
\texttt{\_num\_to\_prefix} matrix is

\[
  W_{n\rightarrow p} \in \mathbb{R}^{
    (\text{N\_PREFIX}\cdot\text{GPT2\_HIDDEN}) \times \text{FUSION\_DIM}}
\]

with $\text{requires\_grad}=\text{False}$ (a frozen random projection
persisted in the checkpoint).

\subsubsection{wte-Stat Matching}

A raw random projection lands far from the GPT-2 token-embedding
distribution. After projecting and reshaping to $\text{N\_PREFIX} \times
\text{GPT2\_HIDDEN}$, the prefix is rescaled to match
$E_{\text{wte}}$ per-tensor statistics:

\[
  \text{prefix} \leftarrow \frac{\text{prefix} - \mu_{\text{prefix}}}
                                {\sigma_{\text{prefix}} + \epsilon}
                          \cdot \sigma_{E_{\text{wte}}}
                          + \mu_{E_{\text{wte}}}
\]

so GPT-2 sees something whose moments resemble a real token embedding
rather than out-of-distribution noise.

\subsubsection{Anchored Decoding}

The driver sentence (\texttt{anchor\_text}) is tokenised, embedded
through GPT-2's $E_{\text{wte}}$, and concatenated \emph{after} the
prefix:

\[
  \text{inputs\_embeds} = [\,\text{prefix},\; E_{\text{wte}}
                          (\text{anchor\_ids})\,]
\]

GPT-2 then continues from real text scaffolding while the prefix
biases the continuation. When no anchor is provided, a single BOS
embedding is used instead.

\textbf{Degenerate-prefix guard}: If $\sigma(\text{prefix}) < 10^{-3}$
the prefix carries no signal and would only push GPT-2 to repeat the
same noise; \texttt{decode()} returns the empty string instead of
calling \texttt{generate()}.

\textbf{LM choice}: GPT-2 is replaced with \texttt{distilgpt2} —
identical tokenizer and $\text{GPT2\_HIDDEN}=768$, roughly half the
weights and $\sim 2\times$ faster. The text is a debug readout, so the
small quality drop is irrelevant.

\section{Training Loop Architecture}

\subsection{Forward Pass}

The forward pass (\texttt{\_forward}) builds a fully differentiable
computation graph:

\[
  \text{model\_pred} = \text{LarkosModel}(x_{\text{temporal}},
    \text{embed\_ctx})
\]

The LarkosModel outputs a vector that is split into an LLM embedding
($\text{llm\_embed}$) and auxiliary dimensions.

\subsubsection{Cross-Attention Path}

The LLM embedding is gated and fused with text context:

\[
  \text{embed\_gate} = \text{EmbedWeightNet}(x_{\text{norm}})
\]

\[
  \text{llm\_embed\_g} = \text{llm\_embed} \odot \text{embed\_gate}
\]

\[
  \text{llm\_embed\_ca} = \text{InputCrossAttention}(x_{\text{norm}},
    \text{llm\_embed\_g})
\]

where $\odot$ denotes element-wise multiplication.

\textbf{Why Cross-Attention}: Instead of concatenating the input and
embedding, cross-attention treats the $D$-dimensional input as $D$ separate
1-D query tokens and the $E$-dimensional embedding as key/value. This is
far more expressive than simple concatenation and allows learned
interaction.

\subsubsection{Cognitive Fusion}

The cross-attention embedding is detached (breaking the gradient graph at
the C boundary) and passed to cognitive fusion:

\[
  \text{fused\_cog\_raw} = \text{cognitive\_fuse}(\text{llm\_embed\_ca},
    \text{neurons}, \text{mem\_state}, \text{default\_weights}, \ldots)
\]

\textbf{Graph Breaking}: The detach at the C boundary is intentional. The
C-side fusion depends on backend neurons and memory state which are
non-differentiable objects. The Python side cannot backprop into them. An
auxiliary loss on the cross-attention embedding (\texttt{aux\_proj})
provides the gradient signal for the modules before the C boundary.

\subsubsection{Frozen Input Windows}

Training targets are refreshed every $\texttt{TARGET\_FREEZE\_INTERVAL}$
epochs. Between refreshes, the target and all fusion-head inputs are held
constant:

\[
  \text{frozen\_input} =
    (\texttt{\_cached\_fused\_cog} \neq \text{None})
    \land (\texttt{\_cached\_driver} \neq \text{None})
    \land (\texttt{\_cached\_graph\_inputs} \neq \text{None})
    \land (\text{epoch} - \text{target\_epoch}) < \text{INTERVAL}
\]

Three caches are maintained:
\begin{itemize}
  \item \texttt{\_cached\_fused\_cog}: the normalised C-side fused vector
  \item \texttt{\_cached\_driver}: the \texttt{llm\_embed\_ca} driver
    embedding
  \item \texttt{\_cached\_graph\_inputs}: the raw GAT inputs
    (node\_features, adj\_mask, edge\_weight) — pinned so the fusion head
    sees a consistent neuron snapshot; the GAT itself still runs
    \texttt{forward\_from\_inputs} in-graph on every step so its
    parameters receive gradient throughout the window
\end{itemize}

All three caches are invalidated together whenever the target is refreshed,
so input and target always move in lockstep.

\textbf{Rationale}: Pinning the inputs and target on the same refresh
schedule prevents the model from chasing a moving input toward a fixed
target, which produces unstable loss. The GAT inputs are pinned at the
C-side boundary, not at the GAT output, so the GAT parameters keep
receiving gradient on every frozen epoch.

\subsection{Neuron Graph Reasoner}

Before the fusion head, the live neuron graph is processed by
\texttt{\_NeuronGraphReasoner}, producing per-neuron token embeddings:

\[
  \mathbf{H}_{\text{graph}} \in
    \mathbb{R}^{1 \times \text{MAX\_NEURONS} \times d_{\text{gat}}}
\]

Each node's 8-dimensional feature vector ($D_{\text{node}} = 8$)
consists of: state, output, layer one-hot (2 dims), tanh-squashed
connection degree, per-neuron state velocity $(\Delta\text{state})$,
output magnitude, and mean outgoing edge weight. A learned per-neuron
embedding $\mathbf{E}_{\text{neuron}} \in
\mathbb{R}^{\text{MAX\_NEURONS} \times d}$ is added so nodes with
identical static features still produce distinct tokens.

Each \texttt{\_GATLayer} computes multi-head attention scores using the
GAT formulation:
\[
  e_{ij} = \text{LeakyReLU}\!\left(
    \mathbf{a}_{\text{src}}^\top W h_i +
    \mathbf{a}_{\text{dst}}^\top W h_j
  \right) \cdot (1 + \tanh(w_{ij}))
\]

where $w_{ij}$ is the C-side edge weight. Non-edges are masked to
$-\infty$ before softmax. Self-loops are always present so an isolated
neuron updates from itself. Two GAT layers ($\texttt{GAT\_HEADS}=2$,
$\texttt{GAT\_LAYERS}=2$) with a residual + post-LayerNorm connect them.

The reasoner exposes two entry points:
\begin{itemize}
  \item \texttt{build\_graph\_inputs(neurons)}: constructs node features,
    adjacency mask, and edge weight matrix from the live neurons dict.
    Also advances the \texttt{\_prev\_states} buffer for state-velocity
    computation.
  \item \texttt{forward\_from\_inputs(nf, am, ew)}: runs the in-graph
    projection and GAT layers on pre-built features. Used by the training
    loop so the GAT input can be cached while the GAT itself remains
    in-graph.
\end{itemize}

\subsection{Fusion Transformer Head}

The \texttt{\_FusionTransformerHead} attends over a mixed token sequence
of length $\text{MAX\_NEURONS} + 3$:

\begin{itemize}
  \item $[\,0 : \text{MAX\_NEURONS})$ — per-neuron tokens from
    \texttt{\_NeuronGraphReasoner}
  \item $[\text{MAX\_NEURONS}]$ — band\_q token (LLM query band,
    $\text{BAND\_Q}=32$ dims projected to $d_{\text{model}}$)
  \item $[\text{MAX\_NEURONS}+1]$ — band\_m token (memory band,
    $\text{BAND\_M}=32$ dims projected to $d_{\text{model}}$)
  \item $[\text{MAX\_NEURONS}+2]$ — driver token (\texttt{llm\_embed\_ca},
    \texttt{INTERNAL\_DIM} dims projected to $d_{\text{model}}$)
\end{itemize}

Token-type embeddings (4 types: graph / q / m / driver) let the encoder
distinguish token kinds without individual positional indices for the 128
graph tokens:

\[
  \text{seq} = \bigl[\mathbf{H}_{\text{graph}} + \mathbf{e}_{\text{graph}},\;
    \mathbf{q}_{\text{tok}} + \mathbf{e}_q,\;
    \mathbf{m}_{\text{tok}} + \mathbf{e}_m,\;
    \mathbf{d}_{\text{tok}} + \mathbf{e}_d\bigr]
\]

A 3-layer encoder ($d_{\text{model}}=64$, 4 heads, $d_{\text{ff}}=128$)
attends over the sequence. Output is pooled by a \emph{learned-query
attention pool}:

\[
  \text{scores} = \frac{\text{enc} \cdot \mathbf{q}_{\text{pool}}}
    {\sqrt{d_{\text{model}}}}
  \quad,\quad
  \text{pooled} = \sum_i \text{softmax}(\text{scores})_i \cdot
    \text{enc}_i
\]

\[
  \text{output} = \text{LinearOut}(\text{pooled})
    \in \mathbb{R}^{\text{MAX\_NEURONS}}
\]

\textbf{Why learned-query pooling}: A softmax over the 131-token sequence
lets the pooling step learn to up-weight informative tokens rather than
giving every token equal voice.

\textbf{Frozen-input}: In-graph dropout and noise are skipped on a
frozen-window epoch so a pinned input is not regularised against a pinned
target.

\subsection{MAML Inner Loop}

The outer-loss term ($L_{\text{outer}}$) is computed against a
\emph{fast clone} adapted to the current task with a few inner SGD steps.
The clone is built once, at \texttt{TrainingLoop} initialisation, via
\texttt{copy.deepcopy(self.model)} — the
\texttt{EmbeddingProjector.\_\_deepcopy\_\_} override ensures the
$\sim 22$M frozen ST parameters are shared, not duplicated.

Each epoch refreshes only the trainable parameters in place:

\[
  \forall p_{\text{main}},p_{\text{fast}}: \;
  p_{\text{main}}.\text{requires\_grad} \Rightarrow
  p_{\text{fast}}.\text{data} \leftarrow p_{\text{main}}.\text{data}
\]

then runs $\text{MAML\_INNER\_STEPS}$ vanilla SGD steps on
$L = \text{MSE}(\text{fast}(x, c), \text{target})$ with
$\text{lr} = \text{MAML\_INNER\_LR}$. The adapted clone is returned
and a single forward pass produces $\text{maml\_pred}$.

\textbf{MAML detach}: \texttt{x\_temporal} is detached before the inner
loop. MAML's inner \texttt{loss.backward()} would otherwise walk back
through the shared temporal encoder graph and free its saved intermediates
before the outer \texttt{\_backward} can use them. Detaching is also
semantically correct: MAML fast-adapts the model, not the input pipeline.
The temporal encoder still receives gradient via \texttt{model\_pred} in
the outer loss.

\textbf{Why a persistent fast clone}: Re-running \texttt{copy.deepcopy}
inside every inner update traverses the entire model tree (including the
frozen MiniLM) per epoch. The persistent clone reduces the per-epoch cost
to the in-place copy of the small trainable subset.

\subsection{Loss Computation}

The loss is a weighted blend of four components:

\[
  L_{\text{total}} = \lambda_1 L_{\text{outer}} + \lambda_2 L_{\text{base}}
    + \lambda_3 L_{\text{pred}} + \lambda_4 L_{\text{aux}}
\]

where:

\[
  L_{\text{outer}} = \text{SmoothL1}(\text{maml\_pred}, \text{target})
\]

\[
  L_{\text{base}} = \text{SmoothL1}(\text{fused}, \text{target})
\]

\[
  L_{\text{pred}} = \text{SmoothL1}(\text{model\_pred}, \text{target})
\]

\[
  L_{\text{aux}} = \text{SmoothL1}(\text{aux\_proj}(\text{llm\_embed\_ca}),
    \text{target})
\]

The weights are:

\[
  \lambda_1 = \text{mc\_blend}
\]

\[
  \lambda_2 = (1 - \text{mc\_blend}) \cdot 0.4
\]

\[
  \lambda_3 = (1 - \text{mc\_blend}) \cdot 0.3
\]

\[
  \lambda_4 = 0.2
\]

where $\text{mc\_blend} \in [0.1, 0.5]$ is derived from Monte Carlo
variance:

\[
  \text{mc\_blend} = \max(0.5 - \sigma^2_{\text{MC}} \cdot 0.5, 0.1)
\]

\textbf{Adaptive Blending}: High MC variance (model uncertainty)
increases the weight on the outer (MAML) loss, which is more stable. Low
variance decreases it. The system adapts loss weights based on model
confidence.

\textbf{SmoothL1 Loss}: Uses $\beta = 0.1$, giving $L_2$ near zero and
$L_1$ far from zero. This prevents saturation unlike Cosine+L1 which
fought itself.

\textbf{Auxiliary Loss}: The aux path carries gradient signal to
cross-attention, embed-weight-net, and text-proj, which sit before the
non-differentiable C boundary.

\subsection{Gradient Balancing}

The embed\_weight\_net branch is often weak. Balancing rescales weak
gradients:

\[
  \text{scale} = \min\left( \frac{\text{norm}_{\text{FT}}}{10 \cdot
    \text{norm}_{\text{embed}}}, 5 \right)
\]

where norms are EMA-smoothed to avoid per-epoch pulses.

\[
  p.\text{grad} \leftarrow p.\text{grad} \cdot \text{scale}
\]

All gradients are then clipped globally:

\[
  \|g\|_2 \leq 1.0
\]

\textbf{Rationale}: Prevents any single module from exploding and ensures
balanced learning across branches.

\subsection{Monte Carlo Sampling}

During forward, the model runs in training mode and takes $T$ stochastic
samples to estimate output variance:

\[
  \text{samples} = [\text{model}(x + \epsilon_1), \ldots,
    \text{model}(x + \epsilon_T)]
\]

where $\epsilon_i \sim \mathcal{N}(0, 0.05^2) \cdot r_{\text{explore}}$
if exploration is above threshold, else zeros (dropout only).

\[
  \sigma^2_{\text{MC}} = \text{Var}(\text{samples})
\]

\textbf{Conditional sample count}: $T$ depends on whether the run is
actively exploring:

\[
  T = \begin{cases}
    \text{MC\_DROPOUT\_T} &
      \text{if } r_{\text{explore}} > \text{EXPLORE\_THRESHOLD} \\
    \max(\lfloor \text{MC\_DROPOUT\_T}/3 \rfloor, 2) &
      \text{otherwise}
  \end{cases}
\]

When the only variance source is dropout, a small $T$ captures the
signal and \texttt{mc\_blend} EMA smooths the rest. The full $T$
fires only when Gaussian exploration noise is actually being injected,
where the variance estimate has to integrate over noisy inputs and
needs more samples to be meaningful. This cuts MC cost roughly $3\times$
on quiet epochs.

\textbf{Rationale}: Uncertainty estimates inform loss weighting. Without
dropout, variance collapses and weights become uniform. Gaussian noise at
high exploration keeps estimates meaningful.

\subsection{Frozen Skip Logic}

On a frozen (input, target) window, once the loss falls below
$\texttt{\_frozen\_skip\_floor}$ ($\approx 0.15$), the optimizer step
is skipped:

\[
  \text{skip} = \text{(in\_frozen\_window)} \land
    (L_{\text{loss}} < 0.15)
\]

\textbf{Rationale}: Continuing to step a memorized pair drives gradients
to $\sim 10^{-4}$ and wastes progress. Skipping prevents the sawtooth
artifact in loss logs.

\section{Emotional and Identity Updates}

After each backward pass, side-effects update the C-side backend:

\subsection{Novelty Derivation}

From loss history:

\[
  \text{novelty} = \min\left(
    \frac{|L_{\text{current}} - \bar{L}|}{\bar{L} + 10^{-8}}, 1.0
  \right)
\]

where $\bar{L}$ is the mean of the loss history. This is blended with the
backend's reflection:

\[
  \text{novelty} = \max(\text{novelty}, \text{reflect\_novelty},
    \text{reflect\_drift})
\]

\subsection{Satisfaction and Emotion}

Satisfaction combines low loss and improving loss:

\[
  \text{sat} = \min(\max(1.0 - L, 0) + \max(L_{\text{prev}} - L, 0), 1.0)
\]

Emotion type and strength are picked based on loss delta and novelty:

\[
  \text{type} = \begin{cases}
    \text{SURPRISE} & \text{if } \text{novelty} > 0.7 \\
    \text{LOVE} & \text{if } L_{\text{prev}} - L > 0.03 \\
    \text{HATE} & \text{if } L - L_{\text{prev}} > 0.03 \\
    \text{SURPRISE} & \text{otherwise (weak)}
  \end{cases}
\]

\[
  \text{strength} = \begin{cases}
    \min(\text{novelty} \cdot 0.6, 0.6) & \text{type} = \text{SURPRISE} \\
    \min(\Delta L \cdot 3.0, 0.6) & \text{type} = \text{LOVE/HATE} \\
    0.1 & \text{deadband}
  \end{cases}
\]

\textbf{Deadband}: The $0.03$ threshold prevents small wiggles from
triggering extreme emotions. The $0.1$ deadband emotion keeps arousal
decaying instead of being pinned high.

\section{Inference: LarkosRunner}

\subsection{Initialization and Loading}

LarkosRunner reconstructs all forward-path modules in the same order as
TrainingLoop so \texttt{load\_checkpoint} can restore onto matching
attribute names without special-casing:

\[
  \texttt{temporal\_encoder},\;
  \texttt{graph\_reasoner},\;
  \texttt{fusion\_transformer},\;
  \texttt{embed\_weight\_net},\;
  \texttt{cross\_attn},\;
  \texttt{text\_proj}
\]

The C-side backend is reloaded with saved memory and network states:

\[
  \text{backend.load\_memory()}, \quad
  \text{backend.load\_network\_states()}
\]

\textbf{Rationale}: Without the trained backend state, cognitive\_fuse
would see empty neurons and memory, and the readout would not reflect
training dynamics. Both halves (PyTorch weights + C-side state) are needed
together; neither is usable alone.

\subsection{Backend Advance}

Before each \texttt{step()} call reads neurons, memory, or input tensor,
\texttt{\_advance\_backend(alpha)} runs the same C-side updates as the
training loop's pre-forward block:

\[
  \texttt{run\_decision\_path},\;
  \texttt{update\_context},\;
  \texttt{process\_neurons},\;
  \texttt{update\_neuron\_states},\;
  \texttt{update\_attractor\_dynamics},\;
  \texttt{update\_affective\_complexity},\;
  \texttt{reshape\_embeddings\_with\_emotion}
\]

Without this, successive \texttt{step()} calls in a multi-step inference
loop read identical C-side state and return identical output.

\textbf{Parameter defaults}: \texttt{alpha} is derived from the live
backend context via \texttt{derive\_alpha\_from\_context} and
\texttt{derive\_alpha\_from\_params}, matching the training loop's
per-epoch derivation. \texttt{mem\_weight\_ratio} defaults to the band
midpoint $\text{BASE} + 0.5 \times \text{RANGE}$, which sits inside the
envelope the model was trained against. Both can be overridden explicitly.

\subsection{Forward Inference}

The runner's \texttt{step} method mirrors \texttt{TrainingLoop.\_forward}
but without:

\begin{itemize}
  \item MC probes (no uncertainty sampling; frozen dropout)
  \item MAML inner loop (single forward pass only)
  \item Auxiliary loss computation (output path only)
\end{itemize}

The temporal encoder and graph reasoner are included in the forward path:

\[
  X_{\text{enc}} = \texttt{temporal\_encoder}(X_{\text{seq}})
  \quad,\quad
  \mathbf{H}_{\text{graph}} = \texttt{graph\_reasoner}(\text{neurons})
\]

The fused cognitive vector is split into bands before passing to the head:

\[
  \text{band\_q},\; \text{band\_m} =
    \texttt{split\_band\_q\_m}(\texttt{normalize}(\text{fused\_cog\_raw}))
\]

All computation runs with \texttt{@torch.no\_grad()}, preventing gradient
accumulation and saving memory.

\subsection{Input and Frozen Window Handling}

Since inference has no training schedule, the frozen-window logic is
disabled:

\[
  \text{frozen\_input} = \text{False}
\]

This mirrors the behavior of a target-refresh epoch in training.

\subsection{Output}

The runner returns a dictionary with:

\begin{itemize}
  \item \texttt{fused}: The final neuron prediction (NumPy array)
  \item \texttt{model\_pred}: Raw LarkosModel output
  \item \texttt{text\_input}: The driver sentence
  \item \texttt{text\_output}: Debug-readout text. Decoded by the text
    bridge (\S\,Text Bridge) from the C-side \texttt{fused\_cog\_raw}
    (FUSION\_DIM) — \emph{not} the smaller MAX\_NEURONS head output —
    and anchored on \texttt{\_current\_text\_input} when one is
    available, so GPT-2 has real text to continue from.
\end{itemize}

\section{Key Design Decisions}

\subsection{Why Detach at the C Boundary}

The C-side cognitive fusion reads non-differentiable backend objects
(neurons, memory, network history). Python cannot build a computation
graph through C code. Detaching prevents error and forces the gradient
signal through the auxiliary loss instead.

\subsection{Why Frozen Input Windows}

Updating the target every epoch while the fusion-head inputs drift (due to
neuron, memory, and driver changes) causes the loss to thrash. Pinning all
inputs and the target on the same refresh schedule prevents chasing a
moving target. The GAT inputs are pinned at the C-side boundary so the
GAT parameters keep receiving gradient on every frozen epoch.

\subsection{Why Separate Projections for Context Tokens}

Each context token (band\_q, band\_m, driver) carries different
information. Separate \texttt{nn.Linear} projections allow each stream to
learn its own mapping into the shared $d_{\text{model}}$ space and let
attention determine their balance against the 128 GAT graph tokens.

\subsection{Why SmoothL1 Loss}

SmoothL1 with $\beta = 0.1$ avoids saturation of cosine losses and the
fighting behavior of Cosine+L1. It is stable and well-behaved for
vector regression.

\subsection{Why Auxiliary Loss}

The cross-attention, embed-weight-net, and text-proj networks sit before
the C boundary. Without a separate loss on their outputs, they would
receive zero gradient (the path to target is broken). The auxiliary loss
regresses the cross-attention embedding to the target dimension and
provides the needed signal.

\section{Scheduler and Optimization}

Training uses:

\begin{itemize}
  \item Optimizer: AdamW with $\text{lr} = \text{BASE\_LR}$,
    $\text{weight\_decay} = 10^{-4}$, $\text{eps} = 10^{-6}$
  \item Scheduler: CosineAnnealingLR with
    $T_{\max} = \text{COSINE\_T\_MAX}$,
    $\eta_{\min} = 10^{-5}$
  \item EMA: Exponential moving average of main model weights (used for
    validation and inference)
\end{itemize}

\textbf{Why AdamW, not Adam}: Adam folds \texttt{weight\_decay} into
the gradient before adaptive scaling, which under-regularises;
AdamW applies it directly to the weights and is the correct
interpretation of $\text{weight\_decay} = 10^{-4}$.

The optimizer covers all trainable modules:
\texttt{model}, \texttt{fusion\_transformer}, \texttt{graph\_reasoner},
\texttt{temporal\_encoder}, \texttt{embed\_weight\_net},
\texttt{cross\_attn}, and \texttt{text\_proj}.

The scheduler steps after each backward pass, ensuring a smooth cosine
decay over the epoch range.

\section{Conclusion}

The Larkos training loop and runner form a unified system:

\begin{itemize}
  \item \textbf{Training}: Full differentiable forward pass,
    multi-component loss, frozen windows, emotional state management
  \item \textbf{Inference}: Fast forward-only pass, loaded weights and
    backend state, no gradient overhead
\end{itemize}

Both share core modules and math but are optimized for their respective
tasks. The design prioritizes stability (frozen windows, gradient
balancing, adaptive loss weights), interpretability (emotional and
identity updates), and efficiency (EMA, frozen sampling).

\end{document}
